A Method for the Spectroscopic Inference of Stellar Parameters

### 2014-07-31 tex/paper.tex
New draft for Sean

% What other people get when they fit this star. How does our inference compare?
\citet{torres12} determined the stellar properties of WASP-14 using several different methods including {\tt MOOG}, {\tt SME}, and {\tt SPC}. While the {\tt MOOG} and {\tt SME} methods are not directly comparable to our technique because they measure individual line strengths, {\tt SPC} utilizes a moderately-sized region of spectrum and a pre-computed synthetic library that enables a direct comparison with our technique. When we make the same assumptions in \citet{torres12}, such as using the same region of spectrum (5050--5360\AA), the same spectral library, and fixing $\log g$ to the value inferred from the exoplanet transit analysis by \citet{joshi09} (amounting to a $\delta$-function prior on $\log g$$=$4.29), our inference of stellar parameters agrees to within 1$\sigma$ (see Table~\ref{table:tests}). Interestingly, when we allow $\log g$ to float, we infer a lower $T_{\rm eff}$ and $\Z$. These results demonstrate the sensitivity to assumptions about $\log g$ as well as the fact that there may exist biases in surface gravity estimates from either spectroscopy, exoplanetary transits, or both.

@@

\subsection{Generic Tests}

In addition to running tests under the same assumptions as historical analyses, we also allow the region kernels to play a role.

%\begin{rant about why this method rocks}
There are many advantages to placing spectroscopic inference within a Bayesian framework. Our technique uses as much as possible of the spectrum. You go to a telescope to take a spectrum, you should use the whole spectrum, not just subsets of the spectrum that are certain well chosen lines. True, these lines have been identified by the accumulated knowledge of experts to show high discriminatory power against the stellar parameters of interest, but there is a disparate gap between these indices and the core fundamental physics knowledge, since calibrations must still be derived against synthetic models. These are coming up with ``fit metrics'' that are similar to distance functions or figures of merit that would be used in approximate Bayesian computing.

These figures of merit make a lot of sense, true. But given that the raw data product from synthetic models from stars is \emph{actually a spectrum} that has been synthesized from a complicated but \emph{physical} model of a star, wouldn't it be better to relate the two through an actual \emph{likelihood function}, a physical description of the process that generates the data? In this case we use a Gaussian process to absorb the computational artifacts/inaccuracy that is a result from a \emph{imperfect} (but still extraordinarily good) stellar model. This allows us to couch the question of what parameters does the star have in a truly \emph{probabilistic} framework, meaning, we can calculate a posterior probability function of the parameters that is grounded in our physical understanding of what a star looks like. This is one of the great things about a \emph{forward-model}.

Ultimately, with the current state of physical models, generating a spectrum from a limited grid, or only being able to tweak certain parameters, the posterior that one can derive is ultimately limited by the accuracy of the spectral synthesis, including any systematic biases that may exist (and they do, given the current spread of performance of stellar libraries).

The main advantage of using a more sophisticated hierarchical likelihood function (Eqn~\ref{eqn:likelihood}) than $\chi^2$, allows the model-data mismatch to become a parameter in the model. That way there is shrinkage between estimates.

%\end{rant}

Note that using global covariance actually lets the parameters float a bit more, and provides some cushion against the tyranny of too many lines. Only so many parameters, can't infinitely tweak the model spectrum for any large region of spectrum. This parameterizes the badness of fit, better models will have a lower covariance, as evidenced by the Kurucz (-14.6) v. PHOENIX (-14.2) global cov amplitudes.

A major advantage of using the MCMC algorithm to explore the multidimensional probability space is that it provides numerical samples in each dimension. Therefore, marginalizing out a parameter (i.e., numerically integrating over a dimension in probability space) is as simple as combining all of the samples in this dimension. This enables us to present a posterior of the stellar parameters $\vT$ (Figure~\ref{fig:stellar_posterior}) which has been marginalized over all of the nuisance hyperparameters. This posterior is the final estimate of the stellar parameters which \emph{incorporates} any inherent uncertainty due to model mismatch (via the covariance hyperparameters) and flux-calibration (via the Chebyshev polynomials).

The benefit of including nuisance parameters and then marginalizing over them is that we can self-consistently model the uncertainty inherent to any spectrum while naturally capturing any degeneracy between the model parameters. If stellar parameters are estimated using a method which ignores these nuisance parameters (e.g., by-eye fitting) and the astronomer arbitrarily inflates the parameter uncertainties to reflect intuition, the degeneracy between parameters is artificially destroyed.

When iteratively fitting a spectrum, we continue to add line covariance kernels until we have covered all of the high amplitude residuals. Typically, we will add line kernels until all residuals greater than three times the amplitude of the global covariance kernel are covered.  As demonstrated by the Gibbs sampler routine, we explicitly fit for the
hyperparameters of the covariance kernels in parallel with the
stellar parameters. While including these extra parameters does increase the
dimensionality and complexity of our model, they do provide several advantages.
\emph{A priori} we do not know which regions of the spectrum are improperly
modeled. The covariance hyperparameters provide a framework to identify these
regions iteratively and in a self-consistent manner.

@@

Another powerful benefit of including the covariance hyperparameters is the result of their ability to quantify and account for model-data mismatch. When a data spectrum is fit with a high-quality spectral library, the pixel residuals are less likely to be correlated, and the covariance structure will have a smaller amplitude and correlation length. This means that the inference on the stellar parameters will be more precise, in the same way that high quality data allows a more precise result. However, if the systematic mismatch between the data and model is large, then the hyperparameters will be larger and the precision of the stellar parameters will naturally inflate to respect the quality of model-data fit. The benefits of using covariance hyperparameters also extend beyond the use case of fitting a single stellar spectrum. If we fit many stars with the same set of synthetic models, we can use the structure of the covariance matrix to improve the models themselves (see \S\ref{subsec:learning}).

\begin{figure*}[!htb]
\begin{center}
  \includegraphics[draft, width=2.5in]{figs/stellar_triangle_Kurucz.png}
  \caption{The posterior probability function of the stellar parameters for WASP-14, an F star, as explored by the MCMC Gibbs sampler. These stellar parameters are marginalized over the Chebyshev ($\Cheb$) and noise parameters ($\Cov$). Figure generated with \texttt{triangle.py} \citep{foreman-mackey14}.
}
\label{fig:stellar_posterior}
\end{center}
\end{figure*}

\begin{figure*}[!htb]
\begin{center}
  \includegraphics[draft,width=0.9\textwidth,height=4in]{figs/spectrum_region.pdf}
  \caption{The spectrum residuals and many random draws from the final covariance matrix showing how the uncertainty in the ``outlier'' lines is identified by the local covariance kernels.}
\label{fig:regions}
\end{center}
\end{figure*}

Talk about giving more reasonable errors using regions, in the biased case.
Show a (Figure~\ref{fig:regions}) of the regions determined, plotted with the increased variance envelope over the residuals.

\begin{itemize}
  \item Spectral parameters can vary by a large margin depending on which spectral library you use ($200~K$ or more). And which technique, which may be a result of choosing certain indices, certain models, or both. Therefore it is in our interest to use as wide a region of spectrum as possible.
  \item Both spectral libraries have stars that they perform better and worse on.
  \item This will be reflected in the increased level of global noise, and number of ``bad'' regions that have been instantiated.
\end{itemize}

Summary end of this section is that we are

1) validating other techniques (WASP14, Kurucz, etc)
2) and two different synthetic libraries give different answers

this can feed into a discussion about tweaking models.

\section{Discussion}
\label{sec:discussion}

Why is it a good idea to use our likelihood function for spectroscopic inference?

By cataloguing the covariance structure of the residuals, especially those generated from strong spectral line mismatch, we collect valuable information about the quality of the synthetic spectra. After fitting many stars, the accumulated knowledge of the data-model mismatch can be used when fitting a new star. The previous structure of the covariance matrix allows us to set priors on what the covariance in certain regions of the spectrum should be, which will speed convergence for this new star. Additionally, after fitting several stars, the average value of the covariance matrix will inform us about the quality of specific spectral lines in the synthetic models.

@@

We can (should?) add an additional level to the hierarchy of
  hyperparameters and add parameters to describe the \emph{population
  characteristics} of poorly modeled spectral lines (mostly the typical width,
  amplitude of these lines).  This will tell us about the frequency and
distribution of spectral modelling errors.

>>>>

% What other people get when they fit this star. How does our inference compare?
\citet{torres12} determined the stellar properties of WASP-14 using several different methods including {\tt MOOG}, {\tt SME}, and {\tt SPC}. While the {\tt MOOG} and {\tt SME} methods are not directly comparable to our technique because they measure individual line strengths, {\tt SPC} utilizes a moderately-sized region of spectrum and a pre-computed synthetic library that enables a direct comparison with our technique. When we make the same assumptions in \citet{torres12}, such as using the same region of spectrum (5050--5360\AA), the same spectral library, and fixing $\log g$ to the value inferred from the exoplanet transit analysis by \citet{joshi09} (amounting to a $\delta$-function prior on $\log g$$=$4.29), our inference of stellar parameters agrees to within 1$\sigma$ (see Table~\ref{table:tests}, WKX). Interestingly, when we allow $\log g$ to float, we infer a lower $T_{\rm eff}$ and $\Z$ (WKX2). These results demonstrate the sensitivity to assumptions about $\log g$ as well as the fact that there may exist biases in surface gravity estimates from either spectroscopy, exoplanetary transits, or both.

@@

\subsection{WASP-14} \label{subsec:wasp}

We use two synthetic libraries to fit WASP-14. The first is the recent publicly available PHOENIX grid (PHOENIX version 16, \citealt{husser13}), which spans wavelengths from 500\AA\ to 5.5$\mu$m and $T_\textrm{eff}$ from 2300-12,000K. The second is a custom ``CfA grid'' synthesized using Kurucz models by Laird and Morse and is the same grid used by {\tt SPC}. This grid is very high quality because it has been hand-calibrated against the Sun and Vega, therefore it performs very well for solar like stars, however it only spans the comparatively narrow wavelength range 5050--5360\AA.

For demonstrative purposes, the first fits we perform use only the Poisson variances from the original data set (WK1 and WP1) and ignore any correlations that might arise in the residual spectrum. This is equivalent to setting $b=1$ and all other amplitudes of $\cov$ to 0, and simplifies the likelihood to the familiar form of $\chi^2$. While the parameter uncertainty returned from these fits appears remarkably good, it is also unrealistic. This small uncertainty results because any likelihood function assumes that the data set is generated from the model\footnote{For example, other ``robust'' likelihood functions like the Student-t distribution assume that there is some likelihood of non-Gaussian ``outlier'' points that will occur. The Gaussian process likelihood that we assume captures the type of ``outliers'' that we might expect from our data.}, meaning that in this paradigm the only way the data set can differ from the model is through the addition of white Poisson noise. While both the PHOENIX and CfA stellar grids are \emph{incredibly good}, this assumption is invalid because the three 3 $\vt_\ast$ parameters are not enough to account for the behavior of thousands of spectral lines that might be in a reasonably large swath of spectrum. Some line strengths might be different from the synthetic model due to slight inaccuracies in the atomic constants or metallicity compositions that are different from solar in a few elements. Thus, in practice, a data spectrum is different from a synthetic model not just by added Poisson noise, but is also \emph{systematically} different in many spectral lines because the model cannot be a perfect representation of reality. Thus, assuming only white Poisson noise for a spectroscopic fit is equivalent to fitting a straight line to a series of data points whose variance is drastically underestimated. A linear least squares fit $\chi^2$ routine might return a small uncertainty estimate, however the slope and intercept of the line might be drastically biased. Therefore, it is clear that they model requires a way to account for this additional source of systematic ``noise.''

While the CfA grid spectra fit well with no obvious features in the residual spectrum, because the PHOENIX spectrum is computed over a very broad spectral range and there is no tuning for specific stars like the Sun, there are often a small amount of  ``outlier'' lines that appear in the residual spectrum (e.g., \citealt{husser13}, Figure 8). For demonstrative purposes, for the next test (WP2) we use the common technique of ``sigma-clipping,'' whereby we manually mask any region of the spectrum which has a more than a 3 sigma residual near the best fit parameters, refit, and then proceed until there are no more outliers. Masking out ``outlier'' spectral lines has the effect of actually changing the final stellar estimate significantly, while also reducing the uncertainty in the stellar parameters. However, several questions arise about the quality of the inference on the stellar parameters. Because sigma-clipping is a procedure, it is impossible to quantify the effect that clipping a given line has on the final parameter estimates \citep{hogg10}. For example, what if a series of lines with $\sim$3 sigma disagreement actually has substantial discriminatory power against the stellar parameters? Clipping these lines would discard important information and lead to an incorrect inference of stellar parameters.

These results motivate the development of a more sophisticated likelihood function which includes a mechanism to account for the systematic mismatch of data spectrum and model spectrum. For the next tests, (WK2 and WP3), we introduce the global covariance kernel (Eqn.~\ref{eqn:global}) to the fit. We can see that the amplitude and scale of the global covariance grows to capture the discrepancy between data spectrum and synthetic spectrum. Including this global covariance broadens the posteriors on the stellar parameters, providing cushion against the tyranny of too many lines, which cause the unreasonably low uncertainty of the previous tests. Therefore, including the global covariance acts as a way to parameterize the goodness or badness of a fit: better spectral models will have a lower covariance, as evidenced by the Kurucz ($10^{-14.6}$) v. PHOENIX ($10^{-14.2}$) global covariance amplitudes. \todo{Figure out a reasonable way to present nuisance parameter values}. It is important to note that this ability to parameterize the quality of fit \emph{self-consistently} is unique to a hierarchical Bayesian model, contrary to systematic error estimates that are added in post-facto (I probably don't want to name names, but e.g., \citealt{torres12, buchhave12}). Because the parameter uncertainty is derived through the model itself, it \emph{preserves correlations} between parameters, like the well known degeneracies between $T_{\rm eff}$, $\log g$, and $\Z$. Arbitrarily inflating the posterior estimates to reflect intuition would artificially destroy any natural degeneracy of parameters.

A major advantage of using the MCMC algorithm to explore the multidimensional probability space is that it provides numerical samples in each dimension. Therefore, marginalizing out a parameter (i.e., numerically integrating over a dimension in probability space) is as simple as combining all of the samples in this dimension. This enables us to present a posterior of the stellar parameters $\vT$ (Figure~\ref{fig:stellar_posterior}) which has been marginalized over all of the nuisance hyperparameters. This posterior is the final estimate of the stellar parameters which incorporates any inherent uncertainty due to flux-calibration (via $\cheb$) and model mismatch (via $\cov$).

\begin{figure*}[!htb]
\begin{center}
  \includegraphics[draft, width=2.5in]{figs/stellar_triangle_Kurucz.png}
  \caption{The posterior probability function of the stellar parameters for WASP-14, an F star, as explored by the MCMC Gibbs sampler. These stellar parameters are marginalized over the Chebyshev ($\cheb$) and noise parameters ($\cov$). Figure generated with \texttt{triangle.py} \citep{foreman-mackey14}.
}
\label{fig:stellar_posterior}
\end{center}
\end{figure*}

For the next tests, (WPX), we introduce the local covariance kernels to track outlier spectral lines. These kernels can map out the regions of the spectrum which are outliers. The glory of these local kernels are that the parameters allow to play a role in data driven spectra. \emph{A priori} we do not know which regions of the spectrum are improperly modeled. The covariance hyperparameters provide a framework to identify these regions iteratively and in a self-consistent manner.  See Figure~\ref{fig:regions} for a cool visualization of what it looks like. \todo{Runs are almost converged. This can feed into a discussion about tweaking models.}

\begin{figure*}[!htb]
\begin{center}
  \includegraphics[draft,width=0.9\textwidth,height=4in]{figs/spectrum_region.pdf}
  \caption{The spectrum residuals and many random draws from the final covariance matrix showing how the uncertainty in the ``outlier'' lines is identified by the local covariance kernels. Plotted with the increased variance envelope over the residuals.
}
\label{fig:regions}
\end{center}
\end{figure*}

In addition to the stark contrast of stellar posteriors derived using only white Poisson noise versus those with the covariance kernels, the results in Table~\ref{table:tests} also reflect the systematic bias that is a function of which spectral library is used as a backend. Depending on your assumptions, there may be systematic differences of 200K or more in $T_{\rm eff}$.

\subsection{Gl~51}
\todo{Tests run using no covariance, covariance, just the lines, just same region of spectrum. Adding local covariance kernels to Na and Ca lines that are underpredicted.}

Because the spectrum of Gl~51 is the mid-infrared, we are unable to use the CfA grid for these comparisons. All Gl~51 tests were made using the PHOENIX grid.

\section{Discussion}
\label{sec:discussion}
%\begin{rant about why this method rocks}
There are many advantages to placing spectroscopic inference within a Bayesian framework. Given the large costs of acquiring a high signal to noise spectrum, it would be advantageous to utilize as much as possible of the spectrum for performing inference. You go to a telescope to take a spectrum, you should use the whole spectrum, not just subsets of the spectrum that are certain well chosen lines. True, these lines have been identified by the accumulated knowledge of experts to show high discriminatory power against the stellar parameters of interest, but there is a gap in the chain of logic between these indices and the core fundamental physics knowledge, since calibrations must still be derived against synthetic models. These ``fit metrics'' (line indices) are similar to distance functions or a figure of merit.

These figures of merit do make a lot of sense, and one can see why they are used, but given that the raw data product from synthetic models from stars is \emph{actually a spectrum} that has been synthesized from a complicated but \emph{physical} model of a star, instead of going through the intermediary step of a figure of merit, wouldn't it be better to relate the two through an actual \emph{likelihood function}, a physical description of the process that generates the data? In this case we use a Gaussian process to absorb the computational artifacts/inaccuracy that is a result from a \emph{imperfect} (but still extraordinarily good) stellar model. This allows us to couch the question of what parameters does the star have in a truly \emph{probabilistic} framework, meaning, we can calculate a posterior probability function of the parameters that is grounded in our physical understanding of what a star looks like. This is one of the great things about a \emph{forward-model}.

The main advantage of using a more sophisticated hierarchical likelihood function (Eqn~\ref{eqn:likelihood}) than $\chi^2$, is that the model-data mismatch becomes a parameter in the model. When the model is of higher quality, the confidence in the stellar posteriors grows and is reflected by a smaller uncertainty. When the model is of lesser quality, the stellar posteriors inflate to reflect the increased uncertainty in the stellar parameters. This hierarchical model also works within a single spectrum when fitting more than one spectral order. If each order has a disparate prediction about the stellar parameters, then the global kernels will have a low amplitude and the final posterior will be an improved estimate. However, if certain orders predict different stellar parameters, the global kernels will inflate to absorb the model-data mismatch and consequently inflate the posteriors on the stellar parameters.
%\end{rant}

@@

The benefits of using covariance hyperparameters also extend beyond the use case of fitting a single stellar spectrum. If we fit many stars with the same set of synthetic models, we can use the structure of the covariance matrix to improve the models themselves. By cataloguing the covariance structure of the residuals, especially those generated from strong spectral line mismatch, we collect valuable information about the quality of the synthetic spectra. After fitting many stars, the accumulated knowledge of the data-model mismatch can be used when fitting a new star. The previous structure of the covariance matrix allows us to set priors on what the covariance in certain regions of the spectrum should be, which will speed convergence for this new star. Additionally, after fitting several stars, the average value of the covariance matrix will inform us about the quality of specific spectral lines in the synthetic models.

@@

\todo{We can (should?) add an additional level to the hierarchy of hyperparameters and add parameters to describe the \emph{population characteristics} of poorly modeled spectral lines (mostly the typical width, amplitude of these lines).  This will tell us about the frequency and distribution of spectral modelling errors. Or we could just speculate about it.}
